---
title: "The State of Data Journalism in Portugal, Italy, Greece and Spain"
subtitle: "What makes data journalism in PIGS different from the rest of Europe?"
format:
  lumo-html: 
    logo: "logo.png"                                     # Optional. Path to a logo displayed above the title
    github-repo: "https://github.com/publico-data/stateofdatajournalism2023-PIGS"        # Optional. Adds a github corner at the top right
    primary-color: "#F14D3E"
    self-contained: true
    is-particlejs-enabled: true                          # Optional: display interactive particles in the document header
    bg-image: ""                                         # Optional: link to an image for the header background
author: Rui Barros e José Volta e Pinto
date: last-modified
---

# Introduction

This document intends to provide a focused analysis of the state of Data Journalism in Portugal, Italy, Greece and Spain. The analysis is based on the [State of Data Journalism 2023 survey](https://ejc.net/state-of-data-journalism-survey-2023) conducted by the European Journalism Centre (EJC).

Although the EJC had conducted the survey since 2021, we opted to only include the data for 2023 since it's the most recent year available. The survey was open between December 7th, 2023 and January 16th, 2024. Participation was encouraged through direct mailing, social media promotion, and asking the DataJournalism.com and European Journalism Centre network of contacts for help in spreading the word. Targeted respondents include full and part time employed data journalists, as well as freelancers; data editors and team leads; trainers, faculty members and educators; students.

Because it's a survey, it's hard to tell how representative the results are of the population it represents. But since there isn't a unified body that states if someone can be considered a data journalist or not, or either a clear dataset of how many data journalists are per country, we believe that the survey is a good proxy to get a sense of the state of data journalism in each country.

The analysis was conducted by PÚBLICO's data jounalism team and the conclusions can only be attributed to them and them alone. European Journalism Centre (EJC) only provided the raw data and did not conduct this analysis.

# Analysis

```{r}
#| label: setup
library(tidyverse)
library(janitor)
library(readxl)
library(DT)
library(ggbeeswarm)
```

## Data cleaning

```{r}
df <- read_excel("stateofdatajournalism2023_open_anonymised_dataset.xlsx")

pigs <- c("Portugal", "Spain", "Italy", "Greece")
rest_eu <- c("Austria", "Belgium", "Bulgaria", "Croatia", "Cyprus", "Czech Republic", "Denmark", "Estonia", "Finland", "France", "Germany", "Hungary", "Ireland", "Latvia", "Lithuania", "Luxembourg", "Malta", "Netherlands", "Poland", "Romania", "Slovakia", "Slovenia", "Sweden", "France")

nrow(df)

df <- df %>%
  filter(work_study_loc %in% c(pigs, rest_eu))

nrow(df)

```
The 2023 survey had 768 respondents, of wich 332 were working or studying in a European country.

```{r}
df %>%
  count(occupation) |> 
  arrange(desc(n)) |>
  datatable()
```

Because the survey wanted to focus on practicing professionals, we decided to only include the respondents that identified themselves as:
 - Full-time employment at a news company/organisation;
 - Full-time freelancer;
 - Part-time freelancer;
 - Editor/ Team lead;
 - Part-time employment at a news company/organisation;


```{r}
df %>%
  filter(occupation == "Other (Please specify)") %>%
  select(occupation, occupation_other) %>%
  datatable()
```

Since one of the options was "Other (Please specify)", we decided to analyse furher so that we could include every single person that we believe qualifies as a data journalist. Some respondents in this category sounded like they were afraid of considering themselves as data journalists, stating things like "I am a full time journalist at a news org but I use data journalism skills sporadically". We decided to include them, but we excluded all answers that worked in Public Relations, Marketing, or Communications.

```{r}
occupations_include <- c("Full-time employment at a news company/organisation","Full-time freelancer","Part-time freelancer","Editor/ Team lead","Part-time employment at a news company/organisation","Other (Please specify)")

occupations_other <- c("Full time - radio journalist","Full time employment at media organisation","I'm full-time employed at a news company, but we don't do a lot of data journalism.","I am a full time journalist at a news org but I use data journalism skills sporadically","I am a freelancer and from time to time I use data journalism","I use data from time to time in my job as editor","Trabajo como periodista sanitaria y ocasionalmente hago periodismo de datos sobre salud","Corresponsal que recurre a veces al periodismo de datos","Freelance e studente", NA)

df <- df %>%
  filter(occupation %in% occupations_include) |> 
  filter(occupation_other %in% occupations_other)
```


```{r}
df <- df %>%
  mutate(pigs = ifelse(work_study_loc %in% pigs, TRUE, FALSE))

df |> 
  group_by(pigs) |>
  count() |> 
  mutate(per = (n/nrow(df))*100) |> 
  datatable()


write_csv(df, "data_pigs.csv")

```

With all these filter applied, we have `r nrow(df)` respondents that we deem to be data journalists in an European country for the purposes of this analysis. Of those, we have `r sum(df$pigs)` (about `r round(sum(df$pigs)/nrow(df)*100, 2)`%) that are from Portugal, Italy, Greece and Spain.

## Global analysis
```{r}
per_country <- df |> 
  group_by(work_study_loc) |>
  count() |> 
  mutate(per = (n/nrow(df))*100) |> 
  arrange(-per)
  
per_country  |>
  datatable()
```

Regarding the number of respondents, Germany is the most represented country (38 journalists), followed by Italy (34), Spain (21) and France(18). Portugal also has 12 respondents and Greece has 6.

```{r}
pop <- read_delim("population_eu.csv", delim = ";")

per_country <- df |> 
  group_by(work_study_loc) |>
  count() |> 
  mutate(per = (n/nrow(df))*100) |> 
  arrange(-per)

per_country <- per_country |>
  left_join(pop, by = c("work_study_loc" = "Country")) |> 
  mutate(per = (n/pop)*1000000)

m <- per_country  |>
  mutate(pigs = ifelse(work_study_loc %in% pigs, TRUE, FALSE))

m <- m |> 
  mutate(pigs = factor(pigs, levels = c(TRUE, FALSE)))

ggplot(m, aes(y = per, x = pigs, colour = pigs, fill = pigs)) +
    geom_beeswarm(cex = 3) +
    theme_minimal() +
    labs(title="Number of data journalists per 1M hab in EU",
          y="Journalists per 1M hab")

```

The absolute numbers can give the impression that PIGS countries have a good number of data journalists (specially compared to other places in Europe). Even when we look at the data per 1M habitants, PIGS countries are not that different from the rest of Europe.

The number of respondents can be although, misleading, since we don't know if the survey didn't got higher response rates in some countries, if the media sector has more workers in each case, etc.

# Main findings

To understand how different PIGS and non-PIGS countries are data journalism wise, we compared the Chi-square test between the two groups. This statistical test is used to determine if there is a significant difference between two groups. The null hypothesis is that there is no difference between the two groups, while the alternative hypothesis is that there is a difference.

Because the Chi-square (χ^2) can show a relationship only by chance, for every relatiosnhip we also calculated the p-value. This can allow us to determine if the relationship verified happens by chance or if it is real. Only values with a p-value lower than 0.05 or closer to it were considered for this analysis.

The current analysis focuses mostly only on the tests where there was a significant difference between the two groups. Other tests were explored but, for the sake of simplicity, we only included the ones that were significant.

## PIGS journalists engage in significantly fewer cross-border investigative projects than their EU counterparts

```{r}
test_table <- table(df$pigs, df$collaboration)
chisq.test(test_table)
```

The data available reveals that only 10 journalists in Portugal, Italy, Greece and Spain collaborate with other journalists from other countries during 2023. The fact that 14% of the respondents in PIGS countries collaborated with other news companies/organisations on a data journalism projects contrasts with the fact that, in the rest of Europe, this value is 32%.

There is a statistically significant difference in the level of collaboration between journalists in PIGS countries and their EU counterparts. With the data suggesting that PIGS journalists engage in significantly fewer cross-border investigative projects compared to other EU journalists.

```{r}
df |>
  group_by(pigs, collaboration) |>
  summarise(n = n()) |>
  filter(collaboration == "Yes") |>
  ggplot(aes(x = pigs, y = n)) +
    geom_bar(stat = "identity", fill = "#F14D3E") +
    theme_minimal() +
    labs(title = "Only 10 journalists in Portugal, Italy, Greece and Spain collaborate with other journalists from other countries during 2023",
         x = "PIGS",
         y = "Number of journalists")

```



## PIGS journalists report significantly worse access to national and local data compared to other EU journalists

Since "access_local" and "access_national" variables use a categorical scale (Very Poor, Poor, Average, Good, and Very Good), we converted them to a numeric scale using the following logic:

```{r}
ordinal_mapping <- c("Very Poor" = 1, "Poor" = 2, "Average" = 3, "Good" = 4, "Very Good" = 5)
ordinal_mapping
```

Because we are using categorical data, we decided to use the Mann-Whitney U test to compare the distributions of the two variables. This test is used to determine if there is a significant difference between two groups. The null hypothesis is that there is no difference between the two groups, while the alternative hypothesis is that there is a difference.


```{r}
df$access_local_numeric <- as.numeric(factor(df$access_local, levels = names(ordinal_mapping), labels = ordinal_mapping))
df$access_national_numeric <- as.numeric(factor(df$access_national, levels = names(ordinal_mapping), labels = ordinal_mapping))

# Separate PIGS and non-PIGS groups
pigs_local <- df$access_local_numeric[df$pigs == TRUE]
non_pigs_local <- df$access_local_numeric[df$pigs == FALSE]

pigs_national <- df$access_national_numeric[df$pigs == TRUE]
non_pigs_national <- df$access_national_numeric[df$pigs == FALSE]

# Perform Mann-Whitney U test
local_mwu <- wilcox.test(pigs_local, non_pigs_local, alternative = "less")
national_mwu <- wilcox.test(pigs_national, non_pigs_national, alternative = "less")

# Print results
print(local_mwu)
print(national_mwu)
```

Both Mann-Whitney U tests showed a significant difference in the distribution of the two variables. 

The results show a clear pattern: journalists in PIGS countries report significantly worse access to both local and national data than their EU counterparts. This means that, compared to journalists in other parts of Europe, they are more likely to struggle when trying to obtain official information.

```{r}
per_quality_acess_national <- df |>
  group_by(pigs, access_national) |>
  count() |> 
  pivot_wider(
    names_from = pigs,
    values_from = n
  ) |> 
  rename(non_pigs = `FALSE`, pigs = `TRUE`) |> 
  filter(!is.na(access_national))

#transform to factor
per_quality_acess_national$access_national <- factor(per_quality_acess_national$access_national, 
  levels = rev(names(ordinal_mapping)), 
  labels = rev(c("Very Poor", "Poor", "Average", "Good", "Very Good")))

per_quality_acess_national <- per_quality_acess_national |> 
  pivot_longer(cols = c(non_pigs, pigs), names_to = "pigs", values_to = "n")

ggplot(per_quality_acess_national, aes(y = pigs, fill = access_national, x = n)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = rev(c("#d73027", "#fc8d59", "#fee090", "#91bfdb", "#4575b4"))) +
  labs(x = element_blank() , y = element_blank(), fill = "Access Quality", title = "Access to national data is significantly worse in PIGS countries") +
  scale_x_continuous(labels = scales::percent) +  #
  theme_minimal()

```

```{r}

per_quality_acess_local <- df |>
  group_by(pigs, access_local) |>
  count() |> 
  pivot_wider(
    names_from = pigs,
    values_from = n
  ) |> 
  rename(non_pigs = `FALSE`, pigs = `TRUE`) |>  
  filter(!is.na(access_local))

#transform to factor
per_quality_acess_local$access_local <- factor(per_quality_acess_local$access_local, 
  levels = rev(names(ordinal_mapping)), 
  labels = rev(c("Very Poor", "Poor", "Average", "Good", "Very Good")))

per_quality_acess_local <- per_quality_acess_local |> 
  pivot_longer(cols = c(non_pigs, pigs), names_to = "pigs", values_to = "n")

ggplot(per_quality_acess_local, aes(y = pigs, fill = access_local, x = n)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = rev(c("#d73027", "#fc8d59", "#fee090", "#91bfdb", "#4575b4"))) +
  labs(x = element_blank() , y = element_blank(), fill = "Access Quality", title = "Access to local data is significantly worse in PIGS countries") +
  scale_x_continuous(labels = scales::percent) +  #
  theme_minimal()

```

In Portugal, Italy, Greece and Spain, 25% of the respondents classified as "Poor" or "Very poor" the acess to data on a national level. A value that contrasts with the rest of Europe, where only one in 10 respondents classified as "Poor" or "Very poor" the acess to data on a national level.

The difference is even more striking when we look at the access to local data. In Portugal, Italy, Greece and Spain, where none of the 72 respondendts classified the acess to local data as very good (in the rest of Europe, 9% classified acess to local data as very good).

## PIGS data journalists are more likely to work in very small teams

Regarding infrastructures and where data journalists work, there are three indicators on the survey that can be analyzed: 
 - Number of journalists in the newsroom;
 - Number of data journalists in a data unit;
 - Job secuirty, where we use two variables:
  - The number of journalists that believe they will be working in data journalism in 5 years;
  - The number of journalists that are working full-time compared to the ones that are part-time.

### Company size

```{r}
company_size_mapping <- c(
    "Just me - I am a freelancer, sole proprietor, etc." = 1,
    "2 to 9 employees" = 2,
    "10 to 49 employees" = 3,
    "50 to 99 employees" = 4,
    "100 to 499 employees" = 5,
    "500+" = 6
)

df$company_size_numeric <- company_size_mapping[df$company_size]
pigs_company_size <- df$company_size_numeric[df$pigs == TRUE & !is.na(df$company_size_numeric)]
non_pigs_company_size <- df$company_size_numeric[df$pigs == FALSE & !is.na(df$company_size_numeric)]

#Mann-Whitney U tests
company_size_mwu <- wilcox.test(pigs_company_size, non_pigs_company_size, alternative = "less")

company_size_mwu
```

There's no significant difference between the distributions of the two variables. This means that there is no difference on the newsroom sizes bewteen southern european countries and the rest of Europe.

### Number of data journalists in a data unit

```{r}
project_team_size_mapping <- c(
    "One, just me" = 1,
    "2-5" = 2,
    "6-9" = 3,
    "10 or more" = 4
)

df$project_team_size_numeric <- project_team_size_mapping[df$project_team_size]

pigs_team_size <- df$project_team_size_numeric[df$pigs == TRUE & !is.na(df$project_team_size_numeric)]
non_pigs_team_size <- df$project_team_size_numeric[df$pigs == FALSE & !is.na(df$project_team_size_numeric)]

team_size_mwu <- wilcox.test(pigs_team_size, non_pigs_team_size, alternative = "less")

team_size_mwu
```
Even though there is no significant difference regarding the size of the newrooms, PIGS journalists are significantly more likely to work alone than their EU counterparts.

```{r}
df |>
  group_by(pigs, project_team_size, project_team_size_numeric) |>
  filter(!is.na(project_team_size_numeric)) |> 
  count() |> 
  pivot_wider(
    names_from = pigs,
    values_from = n
  ) |> 
  rename(non_pigs = `FALSE`, pigs = `TRUE`) |>  
  arrange(desc(project_team_size_numeric)) |>
  ungroup() |>
  select(-project_team_size_numeric) |>
  filter(!is.na(project_team_size)) |> 
  datatable()
```

```{r}
df |>
  group_by(pigs, project_team_size, project_team_size_numeric) |>
  filter(!is.na(project_team_size_numeric)) |> 
  count() |> 
  pivot_wider(
    names_from = pigs,
    values_from = n
  ) |> 
  rename(non_pigs = `FALSE`, pigs = `TRUE`) |>  
  arrange(desc(project_team_size_numeric)) |>
  ungroup() |>
  mutate(non_pigs = non_pigs/sum(non_pigs)*100, pigs = pigs/sum(pigs)*100) |>
  select(-project_team_size_numeric) |>
  filter(!is.na(project_team_size)) |>
  mutate(project_team_size = ifelse(project_team_size == "One, just me", "1", project_team_size)) |>
  mutate(project_team_size = ifelse(project_team_size == "10 or more", "10+", project_team_size)) |>
  mutate(project_team_size = factor(project_team_size, levels = c("1", "2-5", "6-9", "10+"))) |>
  pivot_longer(cols = c(non_pigs, pigs), names_to = "pigs", values_to = "n") |>
  ggplot(aes(x = project_team_size, y = n, fill = pigs)) +
  facet_wrap(~pigs) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#F14D3E", "#4575b4")) +
  labs(x = "Team size", y = "Percentage(%)", fill = "PIGS") +
  theme_minimal()


```

More than half of the respondents from Portugal, Italy, Greece and Spain work alone as data journalists. The syndrome of the lonely-data-journalist also happens in the rest of Europe, but with a much higher percentage (32%). In fact, the most common team size is teams that range from 2 to 5 people.

Combinig the two conclusions above, we are drawn to conclude that even though we can't say there is a significante difference in the size of the newsrooms where data journalists work at, there is one on the size of the data jorunalism teams. This two factors can also explain why Southern European counterparts are more likely to not participate in investigative and collaborative projects. Knowing that a lot of data journalism teams also take a role as a service desk to the newsroom, it can be argued that the typical data journalist in a PIGS country is someone that must take the role of a full team, whithout room to specialization in more than data.


### Job security

It's hard to find a good proxy question to measure the job security of data journalists in the survey. We are drawn to believe that two variables can be used
 - The number of journalists that believe they will be working in data journalism in 5 years;
 - The number of journalists that are working full-time compared to the ones that are part-time.

 Even thought freelancing can be a legitimate option for many professionals that prefer the freedom it gives them, some professionals can also be pushed to it because of the lack of room for them in the newsroom. We took the secound option as a working hypothesis, knowing that it's hard, from the data provided, to know where do respondents belong to.

```{r}
five_years <- table(df$pigs, df$five_years_time)
chi2_test_five_years <- chisq.test(five_years)

chi2_test_five_years 
```

Both in PIGS and non-PIGS, only 10 respondent stated they don't belive they will not be workin in data journalism in 5 years. There's no statistical significant difference between the two groups.

```{r}
full_time_categories <- c(
    "Full-time employment at a news company/organisation",
    "Editor/ Team lead"
)

freelance_part_time_categories <- c(
    "Part-time freelancer",
    "Full-time freelancer",
    "Part-time employment at a news company/organisation",
    "Other (Please specify)"
)

df$occupation_category <- ifelse(df$occupation %in% full_time_categories, "Full-time", "Freelance/Part-time/Other")

occupation <- table(df$pigs, df$occupation_category)

chi2_test_occ <- chisq.test(occupation)
chi2_test_occ

```

The same thing applies to Full time employed vs freelancers. Theres no statistically significant difference between the two groups.


## AI adoption in journalism appears to be faster in PIGS countries

```{r}
x <- df |> 
  filter(ai_usage != "I'm not sure")

contingency_table_usage <- table(x$pigs, x$ai_usage)
chi2_test_usage <- chisq.test(contingency_table_usage)

chi2_test_usage
```
Even though it's right above the common 0.05 threshold to be considered statistically significant, there seams to be a slower adoption of AI in PIGS countries.

```{r}
x |>
  group_by(pigs, ai_usage) |>
  count() |>
  ungroup() |>
  group_by(pigs) |>
  mutate(per = (n/sum(n))*100) |>
  datatable()
```

```{r}
x |>
  group_by(pigs, ai_usage) |>
  count() |>
  ungroup() |>
  group_by(pigs) |>
  mutate(per = (n/sum(n))*100) |>
  ggplot(aes(x = pigs, y = per, fill = ai_usage)) +
  geom_col(position = "stack") +
  scale_fill_manual(
    values = c("#F14D3E", "#4575b4"),  # Only 2 colors provided
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(labels = c("non-PIGS", "PIGS")) +
  labs(x = element_blank(), y = "% of journalists", fill = "AI tools incorporated",
       title = "AI adoption in journalism appears to be faster in PIGS countries"
  ) +
  theme_minimal()
```

Even thought the value is not statistically significant, the data seams to show that journalists in PIGS countries seam to be more likely to be incorporating AI tools in their work when compared to the other eropean data journalists.

```{r}
df |>
  group_by(ai_challenges) |>
  count() |>
  ungroup() |>
  mutate(per = (n/sum(n))*100) |>
  arrange(desc(n)) |>
  datatable()
```


## Gender unbalance 

```{r}
df |> 
  group_by(pigs, gender) |>
  filter(gender == "Female" | gender == "Male") |>
  count() |>
  ungroup() |>
  group_by(pigs) |>
  mutate(per = (n/sum(n))*100) |>
  arrange(desc(n)) |>
  datatable()
```

```{r}
#test for statistical relevance of gender diferences
a <- df |> filter(gender == "Female" | gender == "Male")

chi2_test_gender <- table(a$pigs, a$gender)
chi2_test_gender <- chisq.test(chi2_test_gender)
chi2_test_gender
```

Even though the value is not statistically significant, the data seams to show that journalists in PIGS countries seam to be more equally distributed between male and female when compared to the rest of Europe, where there are more women journalists working in data journalism.

More compelling data about the amount of female journalists working in these countries would be necessary to draw conclusions if the well known gender imbalance is STEM and tech related jobs is also affecting the field of data jounalism in PIGS countries.



